Staged Compilation with Two-Level Type Theory

### 2022-09-20 paper/paper.tex
typos

\begin{itemize}
  \item In Section \ref{sec:tour-of-2ltt} we present an informal syntax of
    two-level type theory, a dependent type theory with support for two-stage
    compilation. We look at basic use cases involving inlining control, partial
    evaluation and fusion optimizations. We also describe feature variations,
    enabling applications in monomorphization and memory layout control.
  \item In Section \ref{sec:formal-2ltt}, we specify the formal notion of models
    and syntax for the variant of 2LTT used in this paper. We mostly work with a
    high-level algebraic presentation, where the syntax is quotiented by
    conversion. However, we also explain how to extract functions
    operating on non-quotiented syntax, by interpreting type-theoretic
    constructions in a setoid model.
  \item
    In Section \ref{sec:staging-algorithm}, we review the standard presheaf
    model of 2LTT \cite[Section~2.5.3]{twolevel}, which lies over the syntactic
    category of the object theory. We show that evaluation in this model yields
    a staging algorithm for closed types and terms. We then extend staging to
    open types and terms which may depend on object-level variables. We show
    stability of staging, which means that staging is surjective up to
    conversion, and we also show that staging strictly preserves all type and
    term formers. Finally, we discuss efficiency and potential practical implementations
    of staging.
  \item
    In Section \ref{sec:soundness} we show soundness of staging, which roughly
    means that the output of staging is convertible to its input. Staging
    together with its stability and soundness can be viewed as a \emph{strong
    conservativity} theorem of 2LTT over the object theory. This means that the
    same constructions can be expressed in the object theory and the
    object-level fragment of 2LTT, up to conversion, and staging witnesses that
    meta-level constructions can be always computed away. This strengthens the
    weak notion of conservativity shown in \cite{capriotti2017models} and
    \cite{twolevel}.
  \item
    In Section \ref{sec:intensional-analysis}, we discuss possible semantic
    interpretations of intensional code analysis, i.e.\ the ability to look into
    the structure of object-level code.
  \item We provide a small standalone implementation of elaboration and staging
    for a two-level type theory, with code samples that expand on the paper
    \cite{staged-demo}.
  \item To our knowledge, this work is the first formalization and implementation of
    staged compilation in the presence of full dependent types, with
    universes and large elimination. In particular, we allow unrestricted
    staging for types, so that types can be computed by metaprograms at compile
    time.
\end{itemize}

@@

Assume $\List_0 : \U_0 \to \U_0$ with $\nil_0 : (A : \U_0) \to \List_0\,A$,
$\cons_0 : (A : \U_0) \to A \to \List_0\,A$ and $\msf{foldr}_0 : (A\,B : \U_0)
\to (A \to B \to B) \to B \to \List_0\,A \to B$. We define a mapping function which
inlines its function argument:
\begin{alignat*}{3}
  & \mit{map} : (A\,B : \Lift\U_0) \to (\Lift\,\spl A \to \Lift\,\spl B)
      \to \Lift(\List_0\,\spl A) \to \Lift(\List_0\,\spl B)\\
  & \mit{map} := \lambda\,A\,B\,f\,\mit{as}.\,
      \qut{\msf{foldr}_0\,
        \spl A\,(\List_0 \spl B)\,
        (\lambda\,a\,\mit{bs}.\, \cons_0\,\spl B\,\spl(f\,\qut{a})\,\mit{bs})\,
        (\nil_0\,\spl B)\,
        \spl as
        }
\end{alignat*}
This $\mit{map}$ function can be defined with quantification over $\Lift \U_0$ but not
over $\U_1$, because $\List_0$ expects type parameters in $\U_0$, and there is
no generic way to convert from $\U_1$ to $\U_0$. Now, assuming
$\blank\!+_0\!\blank : \Nat_0 \to \Nat_0 \to \Nat_0$ and $\mit{ns} :
\List_0\,\Nat_0$, we have the following staging behavior:
\begin{alignat*}{3}
  & &&\spl(\mit{map}\,\qut{\Nat_0}\,\qut{\Nat_0}\,(\lambda\,n.\,\qut{\spl n +_0 10})\,\qut{\mit{ns}})\\
  & =\hspace{0.3em}&&\spl\qut{\msf{foldr}_0\,
        \spl \qut{\Nat_0}\,\spl \qut{\List_0\,\spl\qut{Nat_0}}\\
  & && \hspace{3.5em}(\lambda\,a\,\mit{bs}.\, \cons_0\,\spl\qut{\Nat_0}\,\spl\qut{\spl\qut{a} +_0 10}\,\mit{bs})\,
        (\nil_0\,\spl \qut{\Nat_0})\,
        \spl\qut{\mit{ns}}}\\
  & =\hspace{0.3em} &&\msf{foldr_0}\,\Nat_0\,(\List_0\,\Nat_0)\,\,(\lambda\,a\,bs.\,\cons_0\,(a +_0 10)\,bs)\,(\nil_0\,\Nat_0)\,\mit{ns}
\end{alignat*}

@@

\subsubsection{Fusion}
Fusion optimizations can be viewed as binding time improvement techniques for
general inductive types. The basic idea is that by lambda-encoding an inductive
type, it is brought to a form which can be binding-time improved. For instance,
consider foldr-build fusion for lists \cite{short-cut}, which is employed in GHC
Haskell.  Starting from $\Lift (\List_0\,A)$, we use Böhm-Berarducci encoding
\cite{boehm-berarducci} under the lifting to get
\[ \Lift((L : \U_0) \to (A \to L \to L) \to L \to L) \simeq ((L : \Lift\,U_0) \to (\Lift\,A \to \Lift\,\spl L \to \Lift\,\spl L)). \]
Alternatively, for \emph{stream fusion} \cite{stream-fusion}, we embed $\List\,A$ into the
coinductive colists (i.e.\ the possibly infinite lists), and use a terminal
lambda-encoding. The embedding into the ``larger'' structure enables some staged
optimizations which are otherwise not possible, such as fusion for the
$\msf{zip}$ function. However, the price we pay is that converting
back to lists from colists is not necessarily total.

@@

>>>>

\begin{itemize}
  \item In Section \ref{sec:tour-of-2ltt} we present an informal syntax of
    two-level type theory, a dependent type theory with support for two-stage
    compilation. We look at basic use cases involving inlining control, partial
    evaluation and fusion optimizations. We also describe feature variations,
    enabling applications in monomorphization and memory layout control.
  \item In Section \ref{sec:formal-2ltt}, we specify the formal notion of models
    and syntax for the variant of 2LTT used in this paper. We mostly work with a
    high-level algebraic presentation, where the syntax is quotiented by
    conversion. However, we also explain how to extract functions
    operating on non-quotiented syntax, by interpreting type-theoretic
    constructions in a setoid model.
  \item
    In Section \ref{sec:staging-algorithm}, we review the standard presheaf
    model of 2LTT \cite[Section~2.5.3]{twolevel}, which lies over the syntactic
    category of the object theory. We show that evaluation in this model yields
    a staging algorithm for closed types and terms. We then extend staging to
    open types and terms which may depend on object-level variables. We show
    stability of staging, which means that staging is surjective up to
    conversion, and we also show that staging strictly preserves all type and
    term formers. Finally, we discuss efficiency and potential practical implementations
    of staging.
  \item
    In Section \ref{sec:soundness} we show soundness of staging, which roughly
    means that the output of staging is convertible to its input. Staging
    together with its stability and soundness can be viewed as a \emph{strong
    conservativity} theorem of 2LTT over the object theory. This means that the
    same constructions can be expressed in the object theory and the
    object-level fragment of 2LTT, up to conversion, and staging witnesses that
    meta-level constructions can be always computed away. This strengthens the
    weak notion of conservativity shown in \cite{capriotti2017models} and
    \cite{twolevel}.
  \item
    In Section \ref{sec:intensional-analysis}, we discuss possible semantic
    interpretations of intensional code analysis, i.e.\ the ability to look into
    the structure of object-level code.
  \item We provide a small standalone implementation of elaboration and staging
    for a two-level type theory \cite{staged-demo}.
  \item To our knowledge, this work is the first formalization and implementation of
    staged compilation in the presence of full dependent types, with
    universes and large elimination. In particular, we allow unrestricted
    staging for types, so that types can be computed by metaprograms at compile
    time.
\end{itemize}

@@

Assume $\List_0 : \U_0 \to \U_0$ with $\nil_0 : (A : \U_0) \to \List_0\,A$,
$\cons_0 : (A : \U_0) \to A \to \List_0\,A$ and $\msf{foldr}_0 : (A\,B : \U_0)
\to (A \to B \to B) \to B \to \List_0\,A \to B$. We define a mapping function which
inlines its function argument:
\begin{alignat*}{3}
  & \mit{map} : (A\,B : \Lift\U_0) \to (\Lift\,\spl A \to \Lift\,\spl B)
      \to \Lift(\List_0\,\spl A) \to \Lift(\List_0\,\spl B)\\
  & \mit{map} := \lambda\,A\,B\,f\,\mit{as}.\,
      \qut{\msf{foldr}_0\,
        \spl A\,(\List_0 \spl B)\,
        (\lambda\,a\,\mit{bs}.\, \cons_0\,\spl B\,\spl(f\,\qut{a})\,\mit{bs})\,
        (\nil_0\,\spl B)\,
        \spl as
        }
\end{alignat*}
This $\mit{map}$ function can be defined with quantification over $\Lift \U_0$ but not
over $\U_1$, because $\List_0$ expects type parameters in $\U_0$, and there is
no generic way to convert from $\U_1$ to $\U_0$. Now, assuming
$\blank\!+_0\!\blank : \Nat_0 \to \Nat_0 \to \Nat_0$ and $\mit{ns} :
\List_0\,\Nat_0$, we have the following staging behavior:
\begin{alignat*}{3}
  & &&\spl(\mit{map}\,\qut{\Nat_0}\,\qut{\Nat_0}\,(\lambda\,n.\,\qut{\spl n +_0 10})\,\qut{\mit{ns}})\\
  & =\hspace{0.3em}&&\spl\qut{\msf{foldr}_0\,
        \spl \qut{\Nat_0}\,(\List_0\,\spl\qut{\Nat_0})\\
  & && \hspace{3.5em}(\lambda\,a\,\mit{bs}.\, \cons_0\,\spl\qut{\Nat_0}\,\spl\qut{\spl\qut{a} +_0 10}\,\mit{bs})\,
        (\nil_0\,\spl \qut{\Nat_0})\,
        \spl\qut{\mit{ns}}}\\
  & =\hspace{0.3em} &&\msf{foldr_0}\,\Nat_0\,(\List_0\,\Nat_0)\,\,(\lambda\,a\,bs.\,\cons_0\,\Nat_0\,(a +_0 10)\,bs)\,(\nil_0\,\Nat_0)\,\mit{ns}
\end{alignat*}

@@

\subsubsection{Fusion}
Fusion optimizations can be viewed as binding time improvement techniques for
general inductive types. The basic idea is that by lambda-encoding an inductive
type, it is brought to a form which can be binding-time improved. For instance,
consider foldr-build fusion for lists \cite{short-cut}, which is employed in GHC
Haskell.  Starting from $\Lift (\List_0\,A)$, we use Böhm-Berarducci encoding
\cite{boehm-berarducci} under the lifting to get
\begin{alignat*}{3}
         &\Lift((L : \U_0) \to (A \to L \to L) \to L \to L)\\
  \simeq\,\,&((L : \Lift\,U_0) \to (\Lift\,A \to \Lift\,\spl L \to \Lift\,\spl L) \to \Lift\,\spl L \to \Lift\,\spl L).
\end{alignat*}
Alternatively, for \emph{stream fusion} \cite{stream-fusion}, we embed $\List\,A$ into the
coinductive colists (i.e.\ the possibly infinite lists), and use a terminal
lambda-encoding. The embedding into the ``larger'' structure enables some staged
optimizations which are otherwise not possible, such as fusion for the
$\msf{zip}$ function. However, the price we pay is that converting
back to lists from colists is not necessarily total.

@@

\section{Data Availability Statement}

A standalone prototype implementation is available as \cite{staged-demo}. It
implements elaboration and staging for a 2LTT. It includes a tutorial and code
examples that expand on the contents of Section \ref{sec:tour-of-2ltt}.
